---
title: "RNASeq Tertiary Analysis: Part 4"
author: "Bharat Mishra, Ph.D., Austyn Trull, Lara Ianov, Ph.D."
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Packages loaded globabally

```{r message=FALSE}
# Set the seed so our results are reproducible:
set.seed(2020)

# Required packages

# pathway analysis
library(clusterProfiler)
library(msigdbr)

# We will need them for data handling
library(magrittr)
library(ggrepel)
library(dplyr)
library(tidyverse)
library(readr)

# plotting
library(ggplot2)
library(EnhancedVolcano)
library(ComplexUpset)
library(ComplexHeatmap)
library(RColorBrewer)
```

# Gene Set Enrichment Analysis (GSEA)

# Input DEG list

```{r}
Transplant_vs_Naive_annotated <- read.csv(file = "./results/Transplant_vs_Naive/Transplant_vs_Naive_annotated_DEGlist.csv",
                                          row.names = 1)
```


For the purposes of this workshop, we will execute GSEA using the significant DEGs defined by `padj <= 0.05, log2FoldChange <= -1 | log2FoldChange >= 1` to be consistent with GO analysis cutoff.

(Please refer to lecture website on alternative approach of using the full dataset)

```{r}

```

Quick look at #of DEGs.

```{r}

```

Check for duplicated gene_ids
```{r}

```

## Create log2Fold change ranked vector

```{r}

```

# Get the gene_sets from MSigDB

### MSigDB gene sets


```{r}

```

Look different gene set (gs) collections

```{r}

```


## get Hallmark gene set

```
msigdbr(species, category, subcategory)
```

As an example we will use the `H` category:

```{r}

```

# Run GSEA



```{r warning=FALSE}
# Run GSEA

```

## Explore GSEA results

```{r}
# see head excluding the last collumn for simplicity:

```

## Dot plot

```{r fig.height=8, fig.width=8}

```


```{r fig.height=9, fig.width=8}

```


## Ridge Plot

```{r}

```

## Additional plots

```{r, warning=FALSE, message=FALSE}

```


**Challenge: make the circular network**

```{r fig.height=7, fig.width=10, warning=FALSE,message=FALSE}

```


```{r fig.height=4, fig.width=25}

```

#### Most Positive NES

```{r}

```

```{r fig.height=5, fig.width=6}

```

### Most Negative NES

```{r, message=FALSE}

```

```{r fig.height=5, fig.width=6}

```

Write GSEA results to csv

```{r}
write.csv(gsea_enrichment@result, file = "./results/Transplant_vs_Naive/Transplant_vs_Naive_GSEA_hallmark_padj_fc.csv")
```

## session info

```{r}
sessionInfo()
```